### 2013-10-11 Publications/2013-ISBI14/03-methods.tex
Draft sent to all for ISBI abstract

\subsection{Digital Phantom}
Based on the fiber geometries of the digital phantom 
created for the High angular resolution diffusion imaging 
reconstruction Challenge held in ISBI 2013 
(San Francisco, US), we simulated high resolution 
(0.5mm isotropic) T1 (TE/TR=10/1500ms) and T2 
(TE/TR=90/5000ms) images, as well as a two \gls*{dmri}
images (1.0mm isotropic, $b$=1200, 1 B0 image) with
32 and 64 evenly-distributed directions.
Diffusion is modeled by a restricted and a hindered
compartment, similar to \cite{assaf_composite_2005}.
The phantom includes \gls*{wm} fiber bundles,
\gls*{gm} and \gls*{csf}. Physical properties
(T1/T2 times in ms) used in simulation are
($0.832\pm0.010$ / $79\times10^{-3}\pm0.6\times10^{-3}$)
for \gls*{wm},
($1.331\pm0.013$ / $110\times10^{-3}\pm2.0\times10^{-3}$) for
\gls*{gm} and ($3.5\pm0.1$ / $0.25\pm0.01$) for \gls*{csf}.

@@

In order to create a realistic distortion, we
generated a synthetic phase-difference map 
consistent with the phantom, using the 
tools distributed with the \emph{FSL} package 
\cite{jenkinson_fsl_2012}  and
standard parameters ($\Delta$TE=2.46~ms. for the
field mapping and \emph{effective dwell time} of 
0.77~ms. for the \gls*{epi}).
We defined two regions of smooth dephasing and 
computed the corresponding \gls*{vsm}. Amplitude
of the dephasing maps can be modulated, enabling
us to evaluate the impact of the distortion magnitude.
We generated several \glspl*{vsm} with maximum
displacements raging from 3.80 mm to 7.60 mm. The
latter maximum was chosen at the upper limit of the
typical distortion observed in real datasets.

@@

\begin{figure}[thpb]
   \centering
   \includegraphics[width=0.95\columnwidth]{Fig01-Phantom}
   \caption{Ground-truth digital phantom.
   A) T2-weighted image; B) undistorted \textit{b0} volume;
   C, D) distorted \textit{b0} volumes with opposed phase 
   encoding directions, maximum displacement of 3.80 mm.}
   \label{fig:phantom}
\end{figure}

@@

The framework provides \gls*{dwi} reconstruction and
deterministic tractography,
with \emph{Diffusion Toolkit} \cite{wang_diffusion_2007}
for diffusion tensor imaging (configured with 10 random
seeds per voxel by default), and \emph{MRTrix}
\cite{tournier_mrtrix:_2012} for high angular resolution
diffusion tensor imaging (default parameters set to use
constrained spherical deconvolution, maximum harmonic
order of 6, 150000 desired tracks).
For both options, the seeding regions can be set to use
the distorted-corrected \gls*{wm} mask or the regions used
to generate the ground-truth, this second case mimicking
the usual procedure on real data (where regions are 
typically mapped from the anatomically correct T1).

The evaluation framework is completed by automated 
assessment modules. We evaluated three characteristics
of the correction methods. 
Firstly, we assessed the geometrical correctness
reporting overlap indices of three tissue
probability maps (namely \gls*{csf}, \gls*{wm},
and \gls*{gm}), weighting the average by tissue
volumes.
Secondly, to evaluate the quality of the actual 
signal dropout correction, we studied the 
similarity volume by volume computing the $\ell_1$-norm
correlation index. We report this score
on the \textit{b0} and the average of the remaining 32 
volumes. Thirdly, we studied the impact on the
connectivity matrices reporting the number of
false positives (inexistent connections in the
gold-standard) and false negatives (or connections
lost).

>>>>

\subsection{Digital Phantom}
Based on the fiber geometries of the digital phantom 
created for the \gls*{hardi} 
reconstruction Challenge held in ISBI 2013 
(San Francisco, US), we simulated high resolution 
(0.5mm isotropic) T1 (TE/TR=10/1500ms) and T2 
(TE/TR=90/5000ms) images, as well as two \gls*{dmri}
images (1.0mm isotropic, $b$=1200, 1 B0 image) with
32 and 64 evenly-distributed directions.
Diffusion is modeled by a restricted and a hindered
compartment, similar to \cite{assaf_composite_2005}.
The phantom includes \gls*{wm} fiber bundles,
\gls*{gm} and \gls*{csf}. Physical properties
(T1/T2 times in sec) used in simulation are
($0.832\pm0.010$ / $79\times10^{-3}\pm0.6\times10^{-3}$)
for \gls*{wm},
($1.331\pm0.013$ / $110\times10^{-3}\pm2.0\times10^{-3}$) for
\gls*{gm} and ($3.5\pm0.1$ / $0.25\pm0.01$) for \gls*{csf}.

@@

In order to create a realistic distortion, we
generated a synthetic phase-difference map 
consistent with the phantom, using the 
tools distributed with the \emph{FSL} package 
\cite{jenkinson_fsl_2012}  and
standard parameters ($\Delta$TE=2.46~ms. for the
field mapping and \emph{effective dwell time} of 
0.77~ms. for the \gls*{epi}).
We defined two regions of smooth dephasing and 
computed the corresponding \gls*{vsm}. Amplitude
of the dephasing maps can be modulated, enabling
us to evaluate the impact of the distortion magnitude.
We generated several \glspl*{vsm} with increasing
maximum shifts, ranging 3.80-7.60~mm, covering the
typical range of distortion observed in real datasets.

@@

\begin{figure}[thpb]
   \centering
   \includegraphics[width=0.95\columnwidth]{Fig01-Phantom}
   \caption{Ground-truth digital phantom.
   A) T2-weighted image; B) undistorted \textit{b0} volume;
   C, D) distorted \textit{b0} volumes with opposed phase 
   encoding directions, maximum displacement of 3.80~mm.}
   \label{fig:phantom}
\end{figure}

@@

The framework provides two different options for
\gls*{dwi} reconstruction and deterministic tractography.
Firstly, \emph{Diffusion Toolkit} \cite{wang_diffusion_2007}
for \gls*{dti}, is configured with 10 random
seeds per voxel by default. Secondly, \emph{MRTrix}
\cite{tournier_mrtrix:_2012} for \gls*{hardi}, with
default parameters set to use constrained spherical
deconvolution, maximum harmonic order of 6, and 150000
desired tracks.
For both options, the seeding regions can be set to use
either the distorted-corrected \gls*{wm} mask, or the 
regions used to generate the ground-truth. This second
seeding strategy mimics the usual procedure on real 
data, where regions are typically mapped from the
anatomically correct T1.

The evaluation framework is completed by automated 
assessment modules. We evaluated three characteristics
of the correction methods. 
Firstly, we assessed the geometrical correctness
reporting overlap indices of three tissue
probability maps (namely \gls*{csf}, \gls*{wm},
and \gls*{gm}), weighting the average by tissue
volumes.
Secondly, to evaluate the quality of the actual 
signal dropout correction, we studied the 
similarity volume by volume computing the $\ell_1$-norm
correlation index. We report this score
on the \textit{b0} and the average of the remaining \gls*{dwi}
volumes. Thirdly, we studied the impact on the
connectivity matrices reporting the number of
false positives (inexistent connections in the
gold-standard) and false negatives (or connections
lost).

### 2013-10-11 Publications/2013-ISBI14/04-results.tex
Draft sent to all for ISBI abstract

All the modules described in the previous Section were
conveniently integrated in workflows using \emph{nipype}
\cite{gorgolewski_nipype:_2011}. The choice of this tool
grants the reproducibility of the experiments,
and the evaluation workflow is publicly released.
The results of the proposed experiments are summarized
in \autoref{fig:results} and \autoref{table:results}.
The remaining of this section provides extended descriptions
and interpretation of results.

\begin{figure*}[tpb]
   \centering
   \includegraphics[width=0.9\textwidth]{Fig02-Results}
   \caption{Visual comparison of correction methods results. 
   First row represents
   a coronal section of the \textit{b0} volume. In second row, the outcome
   of tractography after filtering tracks that did not connect two regions.
   Third row shows the associated connectivity matrices. First column shows
   the ground-truth. For the remaining steps, false positives have been marked
   in red color and false negatives in blue. Underlaying in gray color,
   the true positive (correctly detected) connections. }
   \label{fig:results}
\end{figure*}

@@

The three methods under study achieved good results in
terms of geometrical accuracy, as it is shown in the first
row of \autoref{fig:results}. The best numerical scores
were obtained with the \gls*{reb} method (\autoref{table:results}),
that achieved a weighted average overlap of $96.64\%$. Second
best were obtained with \gls*{fmb}. Geometrical correctness
is fundamental in connectivity analysis to spatially locate the
\glspl*{roi} that will define the nodes of the final connectivity
matrix. The clear difference of accuracy between \gls*{reb} and \gls*{fmb}
with respect to \gls*{t2b} infers the latter may not be
an appropriate method for susceptibility correction.

@@

A second workflow investigated the similarity of the recovered
signal with respect to the original (undistorted).
In this second study, \gls*{reb} performed
better than the other two methods, as reported
in \autoref{table:results}. \gls*{reb} scored a $91.0\%$
similarity index for the \textit{b0} volume and an average $97.65\%$
for the remaining 31 \glspl*{dwi}. Again, the second
qualified was \gls*{fmb}, which achieved very close results
for the \glspl*{dwi} ($96.26\%$) but not as good for the \textit{b0}
volume. Visual inspection of the recovered data confirms the 
presented numerical results (\autoref{fig:results}, first row).

@@

Connectivity matrices derived from diffusion tensor imaging 
(using 32 directions \glspl*{dwi}) were completely
hindered by the high complexity of the fibers contained
in the phantom. Tractography discovered successfully only 4 out 
of 27 existing connections from the original (undistorted) data,
with more than 45 false connections.
With the 64-directions dataset (reconstructed with constrained 
spherical deconvolution) we found 23/27 connections,
but we still observed 40 false positives. Therefore, the 
experiments using the 32-directions phantom were discarded. The
results presented in this subsection only refer to the 
64-directions dataset.

Using the strongest distortion (maximum displacement of 7.60 mm),
the number of detected connections
remained the same (23/27), slightly increasing the number of
false positives to 42. The immediate conclusion is that, with highly
complex anatomies (crossing, fanning, etc.) and limited number of
ground-truth connections, connectivity matrices are more sensitive 
to reconstruction and tractography than to distortions.

For the sake of completeness, \autoref{table:results-tractography}
reports the characteristic features of the tractography results
and connectivity matrices, for the fieldmap that caused a maximum
displacement of 3.80 mm. Very similar results were obtained for
7.60 mm, a distortion which is at the limit of the real case. This
results, along with visual inspection (\autoref{fig:results}, second
and third rows) might point to \gls*{fmb} as the best correction method.

@@

Even though all the surveyed methods produced visually 
sound results, our study suggested that \gls*{reb} is the 
most accurate methodology for correction of susceptibility-induced
distortions in \gls*{dwi}. Moreover, \gls*{reb} also achieved
the best results in terms of connectivity, being able to recover
one of two lost connections after distortion. \Gls*{fmb} achieved 
very positive scores. Finally, the \gls*{t2b} method did not achieve
the necessary high-standards to recommend its use. We understand
that specific methods with anisotropic regularization that completely
restrict deformations to the phase-encoding direction would perform
significantly better compared to the standard method included
in this study.
%Finally, \gls*{t2b} confirmed that its correction
%can not be compared, even when using the intensity compensation
%based on the determinant of the deformation field.
%Although track counting has been used as a measurement of success and as a weight of connectivity 
%strength, the number of tracks produced by tractography strongly rely
%on the seeding performed and the smoothing caused by interpolation used
%within correction methodologies.

One relevant limitation of this work is that the \gls*{fmb} is
used both in distortion creation and correction, which could
bias the results. For practical reasons (distortion on the mask, 
signal dropout), the \gls*{fmb} can not achieve a perfect 
geometrical correction as might be
expected if the process was exactly inverse.
Nonetheless, \gls*{reb} performed better than \gls*{fmb}.
A second limitation is found on the impossibility of integrating the
point spread function-based methods for two reasons.
Firstly, the unmet need of a point spread function map consistent
with the synthetic phase-difference map. Secondly, to our knowledge,
there are no implementations of the method readily available.
In practice, this second limitation is narrowed down by the
fact that typical acquisition protocols usually include one or more
datasets among field mapping, reversed-encoding \textit{b0} or T2,
but they do not include the point-spread function mapping.

Future extensions of this work will evaluate
different realizations of the synthetic phase-difference map,
better characterizing the phenomenon. Moreover, the
framework can be enhanced for studying the impact of 
other artifacts as subject's motion or eddy 
currents-derived distortions.

>>>>

All the modules described in the previous section were
conveniently integrated in workflows using \emph{nipype}
\cite{gorgolewski_nipype:_2011}. The choice of this tool
grants the reproducibility of the experiments,
and the evaluation workflow is publicly released.
The results of the proposed experiments are summarized
in \autoref{fig:results} and \autoref{table:results}.
The remaining of this section provides extended descriptions
and interpretation of the results.

\begin{figure*}[tpb]
   \centering
   \includegraphics[width=0.9\textwidth]{Fig02-Results}
   \caption{Visual evaluation of the correction methods.
   First row represents a coronal section of the \textit{b0} volume. 
   In second row, the outcome of tractography after filtering tracks
   that did not connect two regions.
   Third row shows the associated connectivity matrices. All the matrices
   are compared to the ground-truth connections in the phantom. The existing
   connections that were correctly detected at each step are depicted in white
   color. False positives have been marked in red and false negatives in blue.
   Gray color represents the true negatives (non-existent connections correctly detected).}
   \label{fig:results}
\end{figure*}

@@

A summary of quantitative results, computed for the \gls*{dti} 
dataset with maximum shift of 3.80~mm,
is provided in \autoref{table:results}.
The best scores were obtained with the \gls*{reb} 
method, followed by \gls*{fmb}. 
The clear difference of accuracy between \gls*{reb} and \gls*{fmb}
with respect to \gls*{t2b} infers the latter may not be
an appropriate method for susceptibility correction. 
The described performance was constant for \gls*{fmb} 
and \gls*{reb} along the different magnitudes of 
distortion evaluated (maximum phase-encoding direction 
shifts in the range of 3.80-7.60~mm).

@@

A second workflow investigated the similarity of the recovered
signal with respect to the original (undistorted).
In this second study, \gls*{reb} performed
better than the other two methods, as reported
in \autoref{table:results}. \gls*{reb} scored a $91.0\%$
similarity index for the \textit{b0} volume and an average $97.65\%$
for the remaining \glspl*{dwi}. Again, the second
qualified was \gls*{fmb}, which achieved very close results
for the \glspl*{dwi} ($96.26\%$) but not as good for the \textit{b0}
volume. Visual inspection of the recovered data confirms the 
presented quantitative results (\autoref{fig:results}, first row).

@@

Connectivity matrices derived from the \gls*{dti} dataset
were completely hindered by the high complexity of the 
fibers contained
in the phantom. Tractography discovered successfully only 4 out 
of 27 existing connections from the original (undistorted) data,
with more than 45 false connections.
With the \gls*{hardi} dataset (reconstructed with constrained 
spherical deconvolution) we found 23/27 connections,
but we still observed 40 false positives. Therefore, the 
experiments using the \gls*{dti} phantom were discarded. The
results presented in this subsection only refer to the 
\gls*{hardi} dataset.

Using the strongest distortion (maximum shift of 7.60~mm),
the number of detected connections
remained the same (23/27), slightly increasing the number of
false positives to 42. The immediate conclusion is that, with highly
complex anatomies (crossing, fanning, etc.) and limited number of
ground-truth connections, connectivity matrices are more sensitive 
to reconstruction and tractography than to distortions.

For the sake of completeness, \autoref{table:results-tractography}
reports the characteristic features of the tractography results
and connectivity matrices, for the fieldmap that caused a maximum
shift of 3.80~mm. Very similar results were obtained for
7.60~mm. This results, along with visual inspection
(\autoref{fig:results}, second and third rows), might point
to \gls*{fmb} as the best correction method.

@@

Even though all the surveyed methods produced visually 
sound results, our study suggested that \gls*{reb} is the 
most accurate method in terms of geometrical accuracy and
signal dropout recovery. 
The \gls*{t2b} method did not achieve the necessary high-standards
to recommend its use. Nonetheless, we understand that specific
methods with anisotropic regularization that completely
restrict deformations to the phase-encoding direction would perform
significantly better than the standard method presented.
Geometrical correctness of \gls*{dwi} data is fundamental 
in connectivity analysis to spatially locate the
regions which will define the nodes of the final connectivity
matrix.

Regarding tractography, this study revealed that signal
reconstruction and tractography algorithms masked the 
impact of susceptibility distortion on the final connectivity
matrices. Due to this effect, experiments performed on 
the \gls*{dti} phantom were discarded for comparison.
With the \gls*{hardi} dataset, the extracted 
connectivity matrix slightly
changed with distortion. Quantitative differences reported in
\autoref{table:results-tractography} could be more related
to the smoothing derived from interpolation implemented by
each method. Visual results might suggest that \gls*{fmb}
achieved better results.

Although we found connectivity matrices rather invariant with
respect distortion, they are suspected to be significantly
impacted by the susceptibility artifact on real data 
\cite{irfanoglu_effects_2012}.
This hypothesis points to the need of more appropriate 
phantoms with denser connectivity matrices.
State of the art phantoms for tractography usually present a
discrete set of simulated fiber bundles that translate in 
very sparse fiber-endings regions and connectivity matrix.
In the real case, the surface limiting the tracks is densely 
covered by the regions mapped from the anatomical 
dataset, what leads to larger sensitivity with respect 
deformations.

A possible limitation of this work is that \gls*{fmb} is
used for both synthesis and correction of the distortion.
However, practical reasons (i.e. noise, signal dropout)
impede perfect geometrical correction with \gls*{fmb}.
Moreover, \gls*{reb} performed more accurately 
than \gls*{fmb}.

Future extensions of this work will include a 
refined digital phantom. Additional lines for this
work will evaluate different realizations of the 
synthetic phase-difference map, to better characterize
the phenomenon. Also, the framework can be enhanced 
for studying the impact of other artifacts as 
subject's motion or eddy currents-derived distortions.

### 2013-10-11 Publications/2013-ISBI14/05-conclusions.tex
Draft sent to all for ISBI abstract

This paper proposes an evaluation framework to
comprehensively analyze the impact of susceptibility
induced distortion on tractography from \gls*{dmri}
data. Inaccuracy on tractography produce errors
and increased variability in subsequent connectivity 
analyses of the whole-brain. Especially, in regions 
influenced by  significant distortion.
We publicly release the framework and also contribute
with the evaluation of the most widely-used and
readily-available correction methodologies. 
The reverse-encoding method (\gls*{reb}) achieved the
best results correcting for susceptibility-induced 
artifacts. We reported geometrical accuracy, signal
dropout recovery and connectivity matrix detection as
ranking indices.

>>>>

This paper proposes an evaluation framework to
comprehensively analyze the impact of susceptibility
induced distortion on tractography from \gls*{dmri}
data. Inaccuracy on tractography produce errors
and increased variability in subsequent connectivity 
analyses of the whole-brain. This effect is suspected
to be larger in regions influenced by large distortions.
Even though, this study found out that the connectivity
matrix is more largely affected by the tractography method.
We publicly release the framework and also contribute
with the evaluation of the most widely-used and
readily-available correction methodologies. 
The reverse-encoding method (\gls*{reb}) achieved the
best results correcting for susceptibility-induced 
artifacts. We also reported visual evaluation of 
tractography and quantitative results for the 
extracted connectivity matrices.
